---
title: "Assignment I"
author: "Will Dibb"
date: "January 15, 2019"
output:
  html_document: default
  pdf_document: default
---


> Please submit your answers by 5:59 pm on 01/28/2019

#### Section 1: Probability : Total points 50


Q1. Which of the following numbers cannot be probability? Explain why.     

a) -0.0001
b) 0.05
c) 3.415
d) 20%
e) 1

Ans 1.  a) and c). The assumption here is that the question is asking which of the above values can be used to describe a possible probability value, where all probability values exist between 0 and 1. Since -0.0001 is a negative value, 3.415 > 1, and 20% = 0.2, a) and c) cannot describe real probability values. 

Q2. A card is drawn randomly from a deck of ordinary playing cards. The game rules are that you win if the card is a spade or an ace. What is the probability that you will win the game?

Ans 2. ~30.8%. A or B uses the simple statistical rule p(AUB) = p(A)+p(B)-p(AB). In this case, given A is a spade, p(A) is 1 of 4 suits so its probability is 0.25. Given B is an ace, p(B) is 1 of 13 card values so its probability is 0.077. Since there is one card that is both an ace and a spade, p(AB) is 1 in 52 or 0.019. The probability then of a win condition as  described is p(AUB) = 0.25 + 0.077 - 0.019 = 0.308

Q3. An urban hospital has a 20% mortality rate on average for admitted patients. If on a particular day, 17 patients got admitted, what are:

a) the chances that exactly 7 will survive?

b) the chances that at least 15 patients will survive?    

Ans 3. 

a) is ~.04%. The probability of mortality event P{1} = 0.20. The probability of P{0}, is then 0.80. Since this is a binomial distribution (there  are two possible outcomes for n repeated experiments), the probability of 10 mortalities out of 17 events
is P(k successes) = (n,k)(p^k)((1-p)^n-k) and (n,k) = n!/(k!(n-k)!).
P(10 successes) = (17!/(10!(7)!))(0.2^10)(0.8^7) = 0.0004

b) is ~31.0%. The probability that at least 15 patients will survive adds the probabilities of P(k=2), P(k=1), and P(k=0) given 2 or less mortalities for 17 patients. Using the same equation as above:
P(k=2) = (17!/(2!15!))(0.2^2)(0.8^15) = 0.1914
P(k=1) = (17!/(1!16!))(0.2^1)(0.8^16) = 0.0957
P(k=0) = (17!/(0!17!))(0.2^0)(0.8^17) = 0.0225
P(k<=2 and k>=0) = 0.3096

Q4. Let F and G be two events such that P(F) is 0.4, P(G) is 0.8. F and G are independent events. Fill in the remaining elements of the table.

Table    | $G$ | $\bar{G}$ | Marginal
---------|-----|-----------|----------
$F$      |0.32 |   0.08    |   0.4
$\bar{F}$|0.48 |   0.12    |   0.6
Marginal |0.8  |   0.2     |    1

Ans 4. P(FG) given F and G are independent, P(FG) = p(F)p(G) so 0.4*0.8=0.32. 
p(F) = 0.4
p(G) = 0.8
\bar{F} = 0.6 
\bar{G} = 0.2

Q5. Let F and G be two events such that P(F) is 0.2, P(G) is 0.7. Now, the conditional probability P(G|F) is given as 0.4. Fill in the remaining elements of the table.

Table    | $G$ | $\bar{G}$ | Marginal
---------|-----|-----------|----------
$F$      |0.08 |    0.12   | 0.2
$\bar{F}$|0.62 |    0.18   | 0.8
Marginal |0.7  |    0.3    |  1

Ans 5. 
P(F) = 0.2
P(G) = 0.7
\bar{F} = 0.8
\bar{G} = 0.3
P(G|F)*p(F) = p(GF) 
0.4*0.2 = p(GF) = 0.08

Q6. A survey was conducted among 100 patients about smoking status. The following is the sample size split by smoking status (Yes or No) and gender (Male or Female). 


Table    | Smoking (Yes) | Smoking(No) | Total
---------|---------------|-------------|----------
Male     |       19      |        36   | 55
Female   |       13      |        32   | 45
Total    |       32      |        68   | 100 


The probability that a randomly selected patient is a male who smokes is 0.19.  

(a) Fill in all the elements of the table


(b) What is the probability of a randomly selected patient being a female?    
0.45

(c) What is the probability of a randomly selected patient being a smoker?     
0.32

(d)  What is the probability of a randomly selected smoker being a female?    
13/32 = 0.406

#### Section 2: Data Analysis using R: Total points 25


Q1 : Using the dataset provided ("sample_patient_dataset.csv"), the task to build a 2x2 table for the studying the association between age at admission >70 and cardiac arrests. You can either use the sample table given below or build your own. Remember to output both count and % in the table. Be sure to round the % to the nearest integer (e.g, 0.674 will be 67% and 0.675 will be 68%, see notes in Lecture2 on summary statistics as example). Fill in the code in the shaded areas.


Table         | Cardiac Arrests (Yes) | Cardiac Arrests (No) | Total
--------------|-----------------------|----------------------|----------
Age > 70 (%)  |      453 (2%)         |     4728 (20%)       | 5181 (21%)
Age <= 70 (%) |      1672 (7%)        |     17254 (72%)      | 18926 (79%)
Total         |      2125 (9%)        |     21982 (91%)      | 24,107 (100%)


```{r}
### Output 2x2 table for age >70 vs <= 70 and
### percent incidence of cardiac arrests in 
### percent integer value using summary statistics

###clear list, load libraries, read csv input

rm(list = ls())
library(plyr)
library(dplyr)
library(lubridate)
library(tidyr)

d.in <- read.csv("sample_patient_dataset.csv")


# convert dob and admit/discharge dates into date-time format
d.in <- mutate(d.in, dob_form = mdy(d.in$dob),
               hosp_admission_form = mdy(d.in$hosp_admission),
               hosp_discharge_form = mdy(d.in$hosp_discharge))

#read age with correct century in DOB function
correct_century <- function(x, year=1908){
  m <- year(x) %% 100
  year(x) <- ifelse(m > year %% 100, 1900+m, 2000+m)
  x
}

#apply function to formatted DOB
d.in <- d.in %>%
  mutate(dob_form = correct_century(mdy(dob)))

#calculate age
d.in <- mutate(d.in, age_at_admit = interval(dob_form, hosp_admission_form) / dyears(1))

#make cardiac arrest column into factor
d.in <- d.in %>%
  mutate(had_cardiac_arrests = as.factor(had_cardiac_arrests))

#group counts
n_cohort <- d.in %>%
  nrow()

n_over70_ca <- d.in %>%
  filter(age_at_admit > 70 & had_cardiac_arrests == 1) %>% 
  nrow()

n_under70_ca <- d.in %>%
  filter(age_at_admit <= 70 & had_cardiac_arrests == 1) %>%
  nrow()

n_over70_noca <- d.in %>%
  filter(age_at_admit > 70 & had_cardiac_arrests == 0) %>%
  nrow()

n_under70_noca <- d.in %>%
  filter(age_at_admit <= 70 & had_cardiac_arrests == 0) %>%
  nrow()

#group percentages
over70_ca_percent <- round((n_over70_ca/n_cohort)*100)

under70_ca_percent <- round((n_under70_ca/n_cohort)*100)

over70_noca_percent <- round((n_over70_noca/n_cohort)*100)

under70_noca_percent <- round((n_under70_noca/n_cohort)*100)

#concatenated outputs for counts and percentages of each of four groups
#use paste function to convert term by term into character vector result
over70_ca_summary <- paste(n_over70_ca,' (',over70_ca_percent,'%)',sep='')

under70_ca_summary <- paste(n_under70_ca,' (',under70_ca_percent,'%)',sep='')

over70_noca_summary <- paste(n_over70_noca,' (',over70_noca_percent,'%)',sep='')

under70_noca_summary <- paste(n_under70_noca,' (',under70_noca_percent,'%)',sep='')



#create matrix of each of the four groups for both count and percentage
cardiac_age_table <- matrix(c(over70_ca_summary,over70_noca_summary,under70_ca_summary,under70_noca_summary),ncol = 2, byrow = TRUE)

#label columns for cardiac events total and percentages
colnames(cardiac_age_table) <- c("Cardiac Event","No Cardiac Event")
#label rows for age groups
rownames(cardiac_age_table) <- c("Age Over 70", "Age Under 70")

cardiac_age_table

 
```



Q2: Create your own de-identified version of "patient_dataset.csv". Upload your de-identified dataset onto Canvas and write the de-identification code below. You will need to refer to the document "Deidentification.pdf" (on Canvas, look under files -> lectures -> lecture_2).

### Insert code here
```{r}
#clear 
rm(list=ls())
#load libraries
library(plyr)
library(dplyr)
library(lubridate)

#read in csv file data frame
d.in <- read.csv("patient_dataset.csv")

#produce separate table with old and new randomly assigned patient ids
d.newid <- d.in %>%
  select(patient.ids) %>%
  unique()
d.newid$random_id <- sample(nrow(d.newid), replace = FALSE)


#use merge function to add randomly assigned ID column to original dataframe with reference point original patient ids       
d.in <- merge(d.in, d.newid, by = "patient.ids")

#create deidentified data frame
#read in original data frame
d.in <- d.in  %>%
  #lubridate date format 
  mutate(discharge_form = mdy(hosp_discharge), 
         admit_form = mdy(hosp_admission))

#count patient total, then create a random shift function between 1 and 365 to use in dateshifting admit/discharge dates
patient_count <- nrow(d.in)
random_shift <- sample(seq(1,365), size = patient_count, replace = TRUE)

#create dateshifted admit/discharge dates
d.in <- d.in %>%
  mutate(hosp_admit_did = admit_form + ddays(random_shift),
         hosp_discharge_did = discharge_form + ddays(random_shift))

#calculate age of patient before removing dob
d.in <- d.in %>%
  #format dob
  mutate(dob_form = mdy(d.in$dob)) %>%
  #create column with calculated interval
  mutate(temp_interval = interval(dob_form, admit_form)) %>%
  #calculate age as new variable
  mutate(age_at_admit = temp_interval / dyears(1))

  
 #remove PHI factors for admit, discharge, dob
d.in <- d.in %>%
  select(-c(hosp_admission, hosp_discharge, admit_form, discharge_form, dob, dob_form, temp_interval))

  #remove remaining PHI factors
d.in <- d.in %>%
select(-patient.ids, -patient.names, -street_address, -zip_code, -city, -contact_number, -admitting_provider) 
  
#save deidentified dataset as separate csv file
write.csv(d.in, "C:/Users/willd/OneDrive/Documents/Winter 2019 UChicago/32000 Intermediate Analytics/deidentified_patient_dataset.csv", row.names = FALSE, quote = TRUE)


```
